---
title: "Parsing GTEx data (version 7)"
author: Diogo M. Camacho
date: `r format(Sys.Date(), "%Y-%m-%d")`
output: html_notebook
---

Here i will parse the GTEx count data and generate an `R` data frame. This will work on the version 7 of the GTEx data, which has been pre-processed outside of R with a series of `bash` commands, to isolate the different components of the data. I will generate a script to reproduce these transformations and clean up later. The following commands will be incorporated into a single `R` script as a wrapper for the work here.
 
```{r libraries}
library(readr)
library(dplyr)
library(umap)
library(ggplot2)
library(forcats)
library(plotly)
```

# Loading count data
```{r read_counts}
count_data <- readr::read_delim("data/gtex_v7_count_data.gct", 
                                delim = "\t", 
                                col_names = FALSE)
```

# Load gene data
```{r read_genes}
genes <- readr::read_delim(file = "data/gtex_v7_genes.txt", 
                           delim = "\t")
```

# Load samples
```{r load_samples}
samples <- readr::read_delim(file = "data/gtex_v7_sample_ids.txt", 
                             delim = "\t", 
                             col_names = FALSE)
```

# Load annotations
```{r load_annotations}
subject_annotations <- readr::read_delim(file = "data/GTEx_v7_Annotations_SubjectPhenotypesDS.txt", 
                                         delim = "\t")

sample_annotations <- readr::read_delim(file = "data/GTEx_v7_Annotations_SampleAttributesDS.txt", 
                                        delim = "\t")
```

I'm going to count how many samples we have per tissue, and plot the corresponding results.

```{r tissue_representation}
samples <- samples %>% 
  dplyr::mutate(., tissue = sample_annotations$SMTSD[match(X1, sample_annotations$SAMPID)]) %>%
  dplyr::mutate(., type = sample_annotations$SMAFRZE[match(X1, sample_annotations$SAMPID)]) %>%
  dplyr::mutate(., broader_type = sample_annotations$SMTS[match(X1, sample_annotations$SAMPID)])

samples %>%
  dplyr::group_by(., tissue) %>%
  dplyr::count() %>%
  ggplot() + 
  geom_col(aes(y = n, x = forcats::fct_reorder(tissue, n))) +
  coord_flip() + 
  theme_bw() + 
  theme(axis.text = element_text(color = "black", size = 12),
        axis.title = element_blank(),
        axis.ticks.length = unit(0.2, "cm"))

```


# Visualization
I will try and do some `umap` here. Because we have 11k samples, I will sample 25 samples per tissue as representative of the tissue itself and I will only look at the top 10% of genes in terms of abundance, per each of those samples.

```{r reduce_dataset}
utis <- unique(samples$tissue)

sel_samples <- vector(mode = "list", length = length(utis))
for (i in seq(1, length(sel_samples))) {
  x1 <- which(samples$tissue == utis[i])
  if (length(x1) < 25) {
    sel_samples[[i]] <- x1
  } else {
    sel_samples[[i]] <- sample(x = x1, size = 25, replace = FALSE)
  }
}
sel_samples <- unlist(sel_samples)

data_labels <- samples[sel_samples, ]

data_mat <- count_data[, sel_samples]
tmp <- vector(mode = "list", length = ncol(data_mat))
num_els <- ceiling(0.1 * nrow(data_mat))
for (i in seq(1, length(tmp))) {
  tmp[[i]] <- sort(as.matrix(data_mat[, i]), decreasing = TRUE, index.return = TRUE)$ix[1:num_els]
}
tmp <- unique(unlist(tmp))

data_mat <- data_mat[tmp, ]
colnames(data_mat) <- seq(1, ncol(data_mat))

```

```{r umap_viz}
umap_plot <- umap::umap(d = t(data_mat), 
                        method = "naive", 
                        random_state = 123,
                        n_epochs = 100,
                        min_dist = 0.1,
                        n_neighbors = 15)

# build data frame
umap_df <- tibble::tibble(x = umap_plot$layout[, 1],
                          y = umap_plot$layout[, 2],
                          sub_tissue = data_labels$tissue[as.numeric(rownames(umap_plot$layout))],
                          tissue = data_labels$broader_type[as.numeric(rownames(umap_plot$layout))])

p1 <- umap_df %>%
  ggplot() + 
  geom_point(aes(x = x, y = y, color = tissue), alpha = 0.75, size = 2) + 
  theme_bw() +
  theme(axis.title = element_blank(),
        axis.ticks.length = unit(0.2, "cm"),
        axis.text = element_text(size = 12),
        legend.position = "none")

p2 <- umap_df %>%
  ggplot() + 
  geom_point(aes(x = x, y = y, color = sub_tissue), alpha = 0.75, size = 2) + 
  theme_bw() +
  theme(axis.title = element_blank(),
        axis.ticks.length = unit(0.2, "cm"),
        axis.text = element_text(size = 12),
        legend.position = "none")


ggplotly(p1)
ggplotly(p2)



```
